---
title: "Fisheries Modeling Class Notes"
author: "Camila Vargas"
date: "2/6/2018"
output: html_document
---

#Lab 5: Using optim in R

Optimization in R is not as good as in Excel but if you have to do the same thing 1000 time, R has more power.

##Function
You assing a name to a certain function. Everything in the function should depend on the parametes you define in your function. 

Key parts: Name,parameters and the return (what your function retuns)

Example on how to get optimum to run

```{r}
#This is how you would NORMALLY write the function
random.func <- function(x, y, arbitrary) {
   answer <- (x-3.9)^2 + (y-9.65)^2 + 3.6
   
   #code using parameters not being minimized
   print(arbitrary)
   return(answer)
}
random.func(x=3, y=2, arbitrary="Hello world")
#this is the smallest possible value of the objective function
random.func(x=3.9, y=9.65, arbitrary="Hello world")

#all you are doing with arbitrary is to print it out.

```


In order to minimize, you have to indicate what you want to minimize.
Eg: change the value of x and y to minimie the values of random.func
You have to set your function in the right way to be able to use `optim()`

```{r}
#For optim() you need all parameters that will be 
#minimized to be in a single vector
#this returns the objective function
random.func2 <- function(xvec, arbitrary) { #avector as a parameter
   x <- xvec[1] #assing 1 for the first part of the vector and y to the second
   y <- xvec[2]
   answer <- (x-3.9)^2 + (y-9.65)^2 + 3.6
   #code using parameters not being minimized
   print(arbitrary)
   return(answer)
}
x <- random.func2(xvec=c(3,2), arbitrary="Hello world") #x =  62.9325
random.func2(xvec=c(3.9,9.65), arbitrary="Hello world") #when x is 3.9 and y is 9.65 then the answer of my function is 3.6.
```


Using `optim()`

```{r}
optim(par=c(1,1),                   #vector of starting values
      fn=random.func2,              #function to minimize
      method="Nelder-Mead",         #method to use
      arbitrary="Hello, I am HAL.")  #other parameters of fn


#Everytime the function optim() calls the function in order to find the minum it prints Hello, I am HAL (in this case it was 69). Convergence =0 means it got to the right answer.

```

##Lab Exercise 

###Part 1

```{r}
#growth function



exp.growth <- function(N0, r, t0=1975, tn=2010){
  nyears <- tn-t0
  n.vec <- numeric(length = nyears)
  n.vec[1] <- N0
  for(i in 1:(nyears)){
    n.vec[i+1] <-(1+r)*n.vec[i]
  }
  return(n.vec)
}

x <- exp.growth(N0=300, r=0.05)
```


###Part 2: Calculationg NNL

```{r}

get.NLL <- function(N0, r, t0=1975, tn=2010){
  nyears <- tn-t0
  n.vec <-exp.growth(N0=N0, r=r)
  N1981 <- n.vec[7]
  N1988 <- n.vec[17]
  N1998 <- n.vec[24]
  NLL <- (log(435)-log(N1981))^2/(2*0.40^2) + (log(559)-log(N1988))^2/(2*0.47^2)+ (log(2280)-log(N1998))^2/(2*0.36^2)
  
  print(c(N1981, N1988, N1998))
  #print(N1988)
  #print(N1998)
  return(NLL)
}

get.NLL(N0=300, r=0.05)

```

###Part 3: Creating a function for optim

```{r}
minimize.NLL <- function(param.vector, startyear=1975, endyear=2010){
  N0 <- param.vector[1]
  r <- param.vector[2]
  n.vec <-exp.growth(N0=N0, r=r)
  N1981 <- n.vec[7]
  N1988 <- n.vec[17]
  N1998 <- n.vec[24]
  NLL <- (log(435)-log(N1981))^2/(2*0.40^2) + (log(559)-log(N1988))^2/(2*0.47^2)+ (log(2280)-log(N1998))^2/(2*0.36^2)
  
  print(c(N1981, N1988, N1998))
  #print(N1988)
  #print(N1998)
  return(NLL)
}

minimize.NLL(param.vector = c(300, 0.05))

optim(par=c(300, 0.05),                   #vector of starting values
      fn=minimize.NLL,              #function to minimize
      method="Nelder-Mead",          #method to use
      startyear=1975, endyear=2010)       
     


```


#Lab 6: Simulation-estimation method
Simulate using a logistic method and a fleet effort going up and down.

##Part 1: Methods for simulating catches, biomass and effort dynamics

```{r}
logistic.effort <- function(r, K, Bstart, Fstart, nyears, 
                            x, a, sigma, nruns,
                            minF=0, maxF=100, plot=T) {
   Bmat <- matrix(nrow=nruns, ncol=nyears, 
                  dimnames=list(paste("Run",1:nruns), paste("Year",1:nyears)))
   Fmat <- matrix(nrow=nruns, ncol=nyears, 
                  dimnames=list(paste("Run",1:nruns), paste("Year",1:nyears)))
   Cmat <- matrix(nrow=nruns, ncol=nyears, 
                  dimnames=list(paste("Run",1:nruns), paste("Year",1:nyears)))
   Bmat[,1] <- Bstart   
   Fmat[,1] <- Fstart
   umsy <- r/2    #umsy = 1-exp(-Fmsy)
   Bmsy <- K/2    #the Bmsy in a logistic equation
   MSY <- r*K/4
   Bmey <- a*Bmsy   #biomass at maximum economic yield
   for (i in 1:nruns) {
      for (yr in 1:(nyears-1)) {
         Cmat[i,yr] <- Bmat[i,yr]*(1-exp(-Fmat[i,yr]))
         Bmat[i,yr+1] <- Bmat[i,yr] + r*Bmat[i,yr]*(1-Bmat[i,yr]/K) - Cmat[i,yr]
         wt <- rnorm(n=1,mean=0,sd=sigma)     #add some uncertainty
         Fmat[i,yr+1] <- Fmat[i,yr]*((Bmat[i,yr]/(a*Bmsy))^x) * exp(wt)    #new fishing mortality
         Fmat[i,yr+1] <- min(Fmat[i,yr+1],maxF)  #limit fishing mortality
         Fmat[i,yr+1] <- max(Fmat[i,yr+1],minF)  #limit fishing mortality
      } #end of loop over years
      Cmat[i,nyears] <- Bmat[i,nyears]*(1-exp(-Fmat[i,nyears]))
   } #end of loop for nruns
   
   #PLOTTING PART
   if (plot==T) {  #only plot if you need to see the plots, otherwise skip plotting
      par(mfrow=c(3,1), mar=c(0,0,0,0), oma=c(5,5,1,1))
      plot(x=1:nyears,y=Fmat[1,], lwd=2,col="black",type="l",yaxs="i", 
           ylim=c(0,1.05*max(Fmat)), las=1, xaxt="n")
      for (i in 2:nruns) {
         lines(x=1:nyears,y=Fmat[i,], col="black", type="l")
      }
      mtext(side=3,"Fishing mortality", line=-2)
      plot(x=1:nyears,y=Bmat[1,], lwd=2,col="blue",type="l",yaxs="i", 
           ylim=c(0,1.1*K), las=1, xaxt="n")
      for (i in 2:nruns) {
         lines(x=1:nyears,y=Bmat[i,], col="blue", type="l")
      }
      mtext(side=3,"Biomass", line=-2)
      plot(x=1:nyears,y=Cmat[1,], lwd=2,col="red",type="l",yaxs="i", 
           ylim=c(0,1.05*max(Cmat)), las=1)
      for (i in 2:nruns) {
         lines(x=1:nyears,y=Cmat[i,], col="red", type="l")
      }
      mtext(side=3,"Catches", line=-2)
   }
   invisible(list(Bmat=Bmat, Fmat=Fmat, Cmat=Cmat, BdivBmsy=Bmat/Bmsy))
}
#short 5-yr run
temp <- logistic.effort(r=0.2, K=1000, Bstart=1000, Fstart=0.01, 
                        nyears=5, x=1, a=1.8, sigma=0.15, nruns=10,
                        minF=0, maxF=100, plot=T)
temp  #look at what is in there from 5 years of simulation

#examine 100 year runs, these are the low and high fishing scenarios
#scenario 1: a=1.8, x=1, sigma=0.15 (very lightly depleted)
require(manipulate)  #needs manipulate function which only works in Rstudio
manipulate(logistic.effort(r=0.2,K=1000,Bstart,Fstart,nyears,x,a,sigma,nruns),
           Bstart=slider(10,1000, initial=1000),
           Fstart=slider(0.01,0.3, initial=0.01),
           nyears=slider(10,5000, initial=100),
           x=slider(0.01, 5, initial=1),
           a=slider(0.2,1.8, initial=1.8),
           sigma=slider(0,0.5, initial=0.15),
           nruns=slider(1,20,initial=4)   )

#scenario 2: a=0.2, x=0.03, sigma=0.15 (serious depletion)
require(manipulate)  #needs manipulate function which only works in Rstudio
manipulate(logistic.effort(r=0.2,K=1000,Bstart,Fstart,nyears,x,a,sigma,nruns),
           Bstart=slider(10,1000, initial=1000),
           Fstart=slider(0.01,0.3, initial=0.01),
           nyears=slider(10,5000, initial=100),
           x=slider(0.01, 5, initial=0.03),
           a=slider(0.2,1.8, initial=0.2),
           sigma=slider(0,0.5, initial=0.15),
           nruns=slider(1,20,initial=4)   )


```


##Part 2: Kleinsner et al. (2013) method

```{r}
Kleisner.method.final.yr <- function(catches) {
   nyear <- length(catches)
   maxC <- -1  #running variable recording the maximum catch
   maxYr <- -1 #records the year of the maximum catch
   
   #find the year of the highest catch & the highest catch itself
   #If there are two years of equal max, the last
   #  is assumed to be the year of maximum catch
   for (i in 1:nyear) {
      if (catches[i] >= maxC) {  #current year catch is the highest so far
         maxC <- catches[i]      
         maxYr <- i
      }
   }
   
   scenario <- -1  #test to make sure that this works   
   #developing scenario
   if (nyear==maxYr) {
      scenario <- 1
   }
   #fully exploited: make sure it is not scenario 1! 
   if (catches[nyear] > 0.5*maxC & scenario < 0) {      
      scenario <- 2
   }
   #overexploited
   if (catches[nyear] >= 0.1*maxC & catches[nyear] <= 0.5*maxC & 
          nyear > maxYr & scenario < 0) {      
      scenario <- 3
   }
   #collapsed
   if (catches[nyear] < 0.1*maxC & nyear > maxYr & 
          scenario < 0) {      
      scenario <- 4
   }
   
   return(scenario)
}

#TO DO: create test cases that will return 1, 2, 3, 4 by changing
#the catch in the last year
Kleisner.method.final.yr(catches = c(5, 20, 100, 500)) #1

Kleisner.method.final.yr(catches = c(50, 200, 300, 160)) #2

Kleisner.method.final.yr(catches = c(20,500, 300, 70)) #3

Kleisner.method.final.yr(catches = c(200, 400, 30)) #4


```


##Part 3: Applying the Kleisner method to the simulated data

```{r}
apply.Kleisner <- function(a=1.8, x=1, nruns=100) {
  
  temp <- logistic.effort(r=0.2, K=1000, Bstart=1000, Fstart=0.01, 
                        nyears=5, x=x, a=a, sigma=0.15, nruns=nruns,
                        minF=0, maxF=100, plot=F)
    #effort <-  logistic.effort(r=0.2, K=1000, Bstart=1000, Fstart=0.01, nyears=100, x=x, a=a, sigma=0.15, nruns=nruns,minF=0, maxF=100, plot=F)
   
    results <- numeric(length=nruns)
    
    for(i in 1:nruns){
      results[i] <- Kleisner.method.final.yr(catches= temp$Cmat[i,]) #calls the Catch matrix from the logistic.effort equation
    }
   
   return(table(results)) #table function allows me to know how many results I have in each scenario
}

apply.Kleisner()


#x=1, a=1.8

#x=0.03, a=0.2
```

##Part 4: Biomass comparison

```{r}
biomass.method.final.yr <- function(BdivBmsy){
 category <- -1  #test to make sure that this works   
   #developing scenario
   if (BdivBmsy>1.5) {
      category <- 1
   }
   #fully exploited: make sure it is not scenario 1! 
   if ( BdivBmsy >0.5 & BdivBmsy <= 1.5) {   
      category <- 2
   }
   #overexploited
   if (BdivBmsy > 0.2 & BdivBmsy <= 0.5) {      
      category <- 3
   }
   #collapsed
   if (BdivBmsy <= 0.2) {      
      category <- 4
   }
   return(category) 
}
  
   
biomass.method.final.yr(BdivBmsy=c(1,2,1,1.6))  #should be 1
biomass.method.final.yr(BdivBmsy=c(1,2,5,1))  #should be 2
biomass.method.final.yr(BdivBmsy=c(1,2,5,0.4))  #should be 3
biomass.method.final.yr(BdivBmsy=c(1,2,5,0.1)) #should be 4

```


#Lab 7: Marine protected area model in R

###Part 1: Check code for harvesr rate

```{r}
calc.harvest.vec <- function(u.out, ncells, MPA.width)  {
   u.vec <- vector(length=ncells)
   u.vec[] <- u.out       #set every cell to u.out, equivalent to u.vec[1:ncells] <- u.out   
   if (MPA.width > 0) {   #no need to do this if there is no MPA! 
      MPA.begin <- round((ncells-MPA.width)/2)+1  #start cell of MPA
      MPA.end <- MPA.begin + MPA.width -1         #end cell of MPA
      u.vec[MPA.begin:MPA.end] <- 0        
   }
   return(u.vec)
}

calc.harvest.vec(u.out=0.2, ncells = 10, MPA.width = 5)

```

###Part 2: MPA Model

```{r}
MPA.model <- function(r, K, u.out, ncells, nsteps, MPA.width, mrate){
  
  u.vec <- calc.harvest.vec(u.out=u.out, ncells=ncells, MPA.width=MPA.width) #calls the function that defines the harvest in each cell

  pop <- numeric(length = ncells) ##creates a vector for the population
  pop[] <- K #fills in the pop vector with K calues
  
  left.cells <- vector(length = ncells)
  right.cells <- vector(length = ncells)
  
  left.cells[] <- c(ncells, 1:(ncells-1))
  right.cells[] <- c(2:ncells,1)
  
  plot(x=1:ncells, y=pop, xlab="Cell number", lwd=3, ylab="Population size", ylim=c(0, 1.05*max(pop)), type="l", yaxs="i", xaxs="i")
  
  for(i in 1:nsteps){
    leaving <- pop[1:ncells]*2*mrate 
    arriving <- 0.5*leaving[left.cells]+0.5*leaving[right.cells]
    
    surplus= r*pop*(1-pop/K)
    catch = u.vec*pop
    
    #update the population numbers
    pop <- pop + surplus - catch - leaving + arriving    
    
    #plot the population in each cell (uncomment when ready)
    
    lines(x=1:ncells, y=pop, lwd=(nsteps-i+1)/nsteps*3)
  }
  
}

MPA.model(r=0.2, K=1000, u.out=0.4, ncells=21,MPA.width=5, nsteps=10, mrate=0.2)

```

#Lab 8: Bayesian assessment of yelloweye rockfish

Read in files
```{r}
early_CPUE <- read.csv("https://raw.githubusercontent.com/camilavargasp/fisheries-modeling-class/master/8%20early%20CPUE.csv")

late_CPUE <- read.csv("https://raw.githubusercontent.com/camilavargasp/fisheries-modeling-class/master/8%20late%20CPUE.csv")

catch <- read.csv("https://raw.githubusercontent.com/camilavargasp/fisheries-modeling-class/master/8%20catches.csv") 

```



The negative log-likelihood of the model given the data, assumed exponential model with base CV=0.02 plus additional variance. 

```{r}
#getNLL

getNLL <- function(catches, CPUE1, CPUE2, r, K, q1, q2) {
   start.yr <- 1970
   end.yr <- 2011
   years <- start.yr:end.yr
   nyr <- length(years)
   B <- vector(length=nyr)
   B[1] <- K*0.8  #stock depleted to 80% of K in 1970
   NLLprior <- 0
   for (i in 1:(nyr-1)) {
      #logistic model, the max() part ensures it does not go extinct
      B[i+1] = max(0.01, B[i] + r*B[i]*(1-B[i]/K) - catches[i,2]) #keep population at 0.01 if the eqaution goes to 0.
   }
   ##complicated bit where the model figures out if there is a CPUE data point in the particular year of interest or not
   for (i in 1:nyr) {
      current.yr <- years[i]  #find the current year of the model
      #if there is a CPUE1 in that year then add NLL
      if (current.yr %in% CPUE1[,1]) {
         obsCPUE <- CPUE1[CPUE1[,1]==current.yr,2]
         obsCV <- CPUE1[CPUE1[,1]==current.yr,3]
         predCPUE <- B[i]*q1
         #print(paste(obsCPUE, predCPUE, obsCV))
         NLLprior = NLLprior + (log(obsCPUE/predCPUE))^2/(2*obsCV^2)
      }
      #if there is a CPUE2 in that year then add NLL
      if (current.yr %in% CPUE2[,1]) {
         obsCPUE <- CPUE2[CPUE2[,1]==current.yr,2]
         obsCV <- CPUE2[CPUE2[,1]==current.yr,3]
         predCPUE <- B[i]*q2
         NLLprior = NLLprior + (log(obsCPUE/predCPUE))^2/(2*obsCV^2)
      }
   }
   #add -ln(prior) for r, K, q1, q2
   #r~U[0,0.2], K~U[9766,100000], q1~U[0.00001,0.01], q2~U[1e-7, 1e-3]
   NLLprior <- NLLprior -log(1/0.2) +  #prior on r 
      -log(1/(100000-9766)) +   #prior on K
      -log(1/(0.01-0.00001)) +  #prior on q1
      -log(1/(1e-3 - 1e-7))     #prior on q2
   
   return(c(NLLprior,B[nyr]))  #return the NLL and final B
}


```

Run the MCMC chain

```{r}

runMCMC <- function(Cfile, CPUE1file, CPUE2file, 
                    ndraws,
                    rinit, Kinit, q1init, q2init) {
   #read in the data (only want to do this once!)
   catches <- read.csv(file=Cfile, header=T, colClasses="numeric")
   CPUE1 <-   read.csv(file=CPUE1file, header=T, colClasses="numeric")
   CPUE2 <-   read.csv(file=CPUE2file, header=T, colClasses="numeric")
   
   #create matrix to store the sampled draws
   posterior <- matrix(nrow=ndraws, ncol=6, 
                       dimnames=list(paste("Draw",1:ndraws),
                                     c("r","K","q1","q2", "NLLprior", "B2011")))
   
   #calculate the NLLprior for the initial sample
   NLLinit <- getNLL(catches=catches, CPUE1=CPUE1, CPUE2=CPUE2,
                     r=rinit, K=Kinit, q1=q1init, q2=q2init)
   posterior[1,] <- c(rinit, Kinit, q1init, q2init, 
                      NLLinit[1], NLLinit[2])
   naccepted <- 1
   noutbounds <- 1
   
   #r~U[0,0.2], K~U[9766,100000], q1~U[0.00001,0.01], q2~U[1e-7, 1e-3]
   for (i in 2:ndraws) {
      if (i %% 1000 == 0) {  #%% is function returning remainder of i/1000
         print(paste("draw",i,"of",ndraws))
      }
      #the jump function
      rstar <- posterior[i-1,1] + runif(n=1, min=-0.0002, max=0.0002)
      Kstar <- posterior[i-1,2] + runif(n=1, min=-700,max=700)
      q1star <- posterior[i-1,3] + runif(n=1, min=-0.00018,max=+0.00018)
      q2star <- posterior[i-1,4] + runif(n=1, min=-7e-5, max=7e-5)
      
      #check to see if draw outside parameter bounds, if so, stay with previous draw
      if (rstar<=0 || rstar>=0.2 || Kstar<=9766 || Kstar>=100000 || 
             q1star<=0.00001 || q1star>=0.01 || q2star<=1e-7 || q2star>=1e-3) {
         noutbounds <- noutbounds+1
         posterior[i,] <- posterior[i-1,]  #outside bounds, reject draw
      }
      else {
         NLLstar <- getNLL(catches=catches, CPUE1=CPUE1, CPUE2=CPUE2,
                           r=rstar, K=Kstar, q1=q1star, q2=q2star)
         
         #THE KEY MCMC step: whether to accept or reject the new values
         ratio <- exp(posterior[i-1,5] - NLLstar[1])
         
         if (ratio >= 1) {  #better place, accept draw automatically
            posterior[i,] <- c(rstar, Kstar, q1star, q2star, NLLstar[1], NLLstar[2])
            naccepted <- naccepted + 1
         } else {
            if (runif(n=1, min=0, max=1) < ratio) { #accept with probability
               posterior[i,] <- c(rstar, Kstar, q1star, q2star, NLLstar[1], NLLstar[2])
               naccepted <- naccepted + 1
            }
            else {
               posterior[i,] <- posterior[i-1,]  #reject and keep last draw
            }
         }
      }
   }
   #plot the posterior
   par(mfrow=c(3,2), oma=c(1,1,1,1), mar=c(2,4,1,1))
   plot(posterior[,1], type="l", ylab="r")
   plot(posterior[,2], type="l", ylab="K")
   plot(posterior[,3], type="l", ylab="q1")
   plot(posterior[,4], type="l", ylab="q2")
   plot(posterior[,5], type="l", ylab="NLL+NLprior")
   plot(posterior[,6]/posterior[,2], type="l", ylab="B2011 / K")  #finalB/K = how depleted stock is
   
   print("accepted divided by draws")
   print(naccepted/ndraws)
   print("draws outside bounds")
   print(noutbounds/ndraws)
   
   #return the results
   return(posterior)
}
x <- runMCMC(Cfile="8 catches.csv", CPUE1file="8 early CPUE.csv", 
             CPUE2file="8 late CPUE.csv", ndraws=1000,
             rinit=0.05, Kinit=12000, q1init=0.002, q2init=1e-4)
acf(x[,1])[1]
acf(x[,2])[1]
acf(x[,3])[1]
acf(x[,4])[1]
Y <- x[seq(20100,to=100000,by=500),]  #burn in and thinning
acf(Y[,1])[1]
acf(Y[,2])[1]
acf(Y[,3])[1]
acf(Y[,4])[1]

x <- runMCMC(Cfile="8 catches.csv", CPUE1file="8 early CPUE.csv", 
             CPUE2file="8 late CPUE.csv", ndraws=10000,
             rinit=0.05, Kinit=12000, q1init=0.002, q2init=1e-4)

```

When this number is big then your jumps are very small so you want to make your jamps bigger. This meand increasing the range where you can jump (when assigning different min and max values in the jump function)

Eg:
"accepted divided by draws"
[1] 0.755 

We are aiming an accepting rate between 20-30%

Most important assumption is that there is the same probability to jump everywhere. Therefore min and max have to have the same absolute values. 


Burn-in and thinin

```{r}
burned <- x[2001:10000]

#thinned <- burned[seq(0.8*ndraws/1000,0.8^ndraws, 0.8*ndraws/1000)]

thinned <- burned[seq(8,8000,8)]


acf(thinned)[1]
acf(x[,1])[1]
acf(x[,2])[1]
acf(x[,3])[1]
acf(x[,4])[1]


```


The longer the chain you run the more sample you are going to end up whent you burn and thin, that are not auto-correlated to each other.

1000 is the minimum you want.








